\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2018-01-19 eg6-letters/NIM/EG6_RTPC_NIM.tex
update figures 6, 7 and 8 and adding the source files in the directory "fig"
editing typos, changing the layout of the figures to fit and be ordered with their numbers

\begin{keywords}
Time projection chamber, gas electron multipliers, alpha particles, DVCS, 
Nuclear physics
\end{keywords}

@@

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nucleus) necessitates an upgrade of the photon detection system.  
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 experiment 
in 2005~\cite{Girod:2007aa}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the CLAS detector and the IC from the 
large flux of low energy M{\o}ller electrons, a 5~T solenoid magnet was placed 
around the target to shield the detectors.  To detect recoiling $\alpha$ 
particles from the coherent DVCS on Helium, a new radial time projection 
chamber (RTPC) was developed to track low energy nuclear fragments. The 
solenoid field was used to bend tracks and measure momentum of particles in the 
RTPC. The CLAS detector supplemented with both IC and RTPC was used during a 
three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from 
coherent DVCS have an average momentum around 300~MeV/c (12~MeV kinetic 
energy). Such low energy $\alpha$ particles are stopped very rapidly, so the 
RTPC was designed to be as close as possible to the target and fit inside the 
230~mm diameter shell and cryostat wall of the solenoid magnet bore of the 
CLAS.

@@

The main structural improvement compared to the 2005 RTPC~\cite{Fenker:2008zz} 
was to obtain a better support structure for the GEM foils. In the previous version,
the RTPC was essencially composed of two independent half-cylinder separated
by their own structures. In this design, the installation of the GEMs was not very practical
and wrinkles were visibles on the GEM surfaces. To allow for a better installation 
and, at the same time, keep the mechanical structure out of the drift volume, 
the present design is based on self supporting GEM cylinder. We 
realized these cylinders by using fiber glass rings glued to 
each ends of the GEM foils in order to 
instal them independently in the RTPC after gluing and soldering operations. The 
rigidity of the GEM foils was enough for the structure to be self-supporting 
and only the upstream end of the cylinder was fixed to the main mechanical 
support structure. This design only left a light fiberglass ring in the 
downstream end, reducing to a minimum secondary interactions.

@@

%To be rewritten!!!
After the ionization, the released electrons drift to the cylindrical detection 
plane under the effect of the electric field. The electrons released close to 
the cathode take the most time to reach the readout pads, the cylindrical 
symmetry indicates that this maximum drift distance is the same for all tracks.  
We illustrate this in Figure~\ref{fig:RTPC_signals}, where we depict a typical 
$^{4}$He track. Due to the non-perfect experimental conditions, in particular 
possible contamination of our gas mixture, the drift speed changed during the 
three months long experimental run\footnote{We had to increase the gas flow 
during the experiment due to a small leak in the RTPC, which concurred to 
the large shift of drift time observed around run number 61600. While our 
gas system was kept slightly over atmospheric pressure to limit 
contamination from air or other external gases, it is likely that this leak was 
the source of modification of the drift speed.}. This can be calibrated by 
finding drift speed as function of experimental run-number. By measuring a 
reference time, we can therefore infer the drift speed of the electrons in the 
RTPC, by: $\textnormal{drift time} = \frac{\textnormal{drift path 
length}}{\textnormal{drift speed}}$.

\begin{figure}[tb]
\centering
\includegraphics[{trim={0 12cm 0 0},clip, scale=0.4}]{RTPC_1_4.pdf}
\caption[]{A schematic drawing of a $^{4}$He track (in green) traversing the 
drift region, with the drift paths followed by the electrons (in black). } 
\label{fig:RTPC_signals}
\end{figure}

Figure~\ref{fig:TDC_profile} presents the time profile of hits from identified 
tracks in one z-bin along the detector in one experimental run. We can clearly 
observe the dropping edge expected from geometrical considerations. In order to 
avoid the statistical fluctuations on finding the maximum drift time 
($T_{max}$) from the time profile, we define a more stable value, i.e., 
$T_{max/2}$, at which the dropping edge passes half the maximum number of hits 
in the histogram. This value was measured in bins along the 200~mm RTPC's 
length to take into account variations in the electric and magnetic field in 
the RTPC (see Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{hits_time_profile.pdf}
\caption[]{Time distribution of the hits in one z-bin along the RTPC in one 
experimental run. Trigger time is defined as T=15, the time unit is the ALTRO 
sampling time of 100~ns.} \label{fig:TDC_profile}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{RunNumber_61452_TDCmax_Zslice.pdf}
\caption{Maximum time of collected hits as a function of the track
              position on the $z$-axis for one experimental run. } 
\label{fig:RunNumber_61551_TDCmax_Zslice}
\end{figure}

Figure~\ref{fig:Drift_run_number_1} shows the $T_{max/2}$ values for individual 
runs (approximately 2 hours long). We observe significant change in the drift 
speed before and after run 61600, while variations within these periods are 
around 2\%.

\begin{figure*}[tb]
\centering
\includegraphics[width=15.5cm]{Drift_run_number.pdf}
\caption{$T_{max/2}$ versus the experimental run numbers.} 
\label{fig:Drift_run_number_1}
\end{figure*}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.46]{FitResult_p2_11.pdf}
\caption{$\Delta \phi$ versus T distribution for tracks
from one bin in longitudinal position along the RTPC. The black line represents 
the final drift paths in this bin.}
\label{fig:DELTA_PHI_TDC}
\end{figure}

@@

@@

We then compared this simulation to experiment on an event by event basis as 
shown in Figure~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated 
as the ratio of the measured ADCs to the simulated deposited energy. This 
provided our initial set of gains for each pads and it was applied to the 
experimental data. However, after this calibration some pads recorded lower ADCs 
than expected. So, we added a correction factor obtained by comparing  
the energy deposit on a given pad to its neighbours. To do so, we compared, for 
each track, the energy deposit on a pad to the energy deposit in the whole track. 
The gain correction factors are extracted by averaging this ratio for a sample 
of elastic events. Final results are shown in Figure~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see the band for $^4$He in its expected position\footnote{The lower peak 
   in the left module of the RTPC comes from an unidentified problem in this half 
   of the detector that concerns 7\% of all the elastic events. 
   Our best guess is that these events are linked with a high voltage supply 
   issue, that would sometimes bring the GEM gain down. These events pass all 
   the elastic requirements and we found nothing that differenciates them 
   from other events but their low recorded ADCs. For 
   the calibration procedures, these events were therefore excluded.}.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Figure~\ref{fig:noise} for a particularly noisy 
channel. Its amplitude is similar to those of real tracks. About 18\% of the 
readout channels exhibit large contributions from this noise.  
Due to its unique time-energy correlation, we use a patern
recognition algorithm and discard the hits coming from the channels that display 
this behavior on an event by event basis. The result of the
procedure is illustrated in the bottom panel of Figure~\ref{fig:noise}.

@@

\subsection{Energy Loss Corrections}\label{sec_eloss}
Energy loss between the target and drift region was significant in the RTPC and 
necessitated a correction for optimal momentum reconstruction at the primary 
interaction vertex.  The dominant loss was in the 27~\textmu{}m thick Kapton target 
wall, with significant contributions also from the pressurized target gas and 
the foils before reaching the drift region.  Corrections were developed based 
on GEANT4 simulations with the full RTPC geometry and parameterized in terms of 
recoil curvature in the drift region and polar angle, separetely for all recoil 
hypotheses ($p$, $d$, $^3$H, $^3$He, $^4$He).  At our average coherent $^4$He 
DVCS kinematics, energy losses were about 5 MeV, while for $e-^4$He elastic 
scattering losses were about 3 MeV, which corresponds to momentum corrections 
of 25\% and 15\%, respectively.

>>>>

\begin{keywords}
Time projection chamber, gas electron multipliers, alpha particles, DVCS, 
Nuclear physics.
\end{keywords}

@@

For certain experiments the base CLAS system was complimented with ancillary 
detectors. For example, the measurement of the Deeply Virtual Compton 
Scattering (DVCS) process ($eH \rightarrow e' H' \gamma$, where $H$ is a 
nucleon or nucleus) necessitates an upgrade of the photon detection system.  
Indeed, with a 6 GeV electron beam, the majority of DVCS photons are produced 
at very forward angles, where the acceptance of the EC was poor. To extend the 
detection range, an inner calorimeter (IC) was built for the E01-113 experiment 
in 2005~\cite{Girod:2007aa}. The IC was constructed from 424 lead-tungstate 
(PbWO$_{4}$) crystals, covering polar angles between 5$^{\circ}$ and 
15$^{\circ}$~\cite{Hyon-suk}. To protect the CLAS detector and the IC from the 
large flux of the low energy M{\o}ller electrons, a 5~T solenoid magnet was 
placed around the target to shield the detectors.  To detect recoiling $\alpha$ 
particles from the coherent DVCS on Helium, a new radial time projection 
chamber (RTPC) was developed to track low energy nuclear fragments. The 
solenoid field was used to bend tracks and measure momentum of particles in the 
RTPC. The CLAS detector supplemented with both IC and RTPC was used during a 
three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from 
coherent DVCS have an average momentum around 300~MeV/c (12~MeV kinetic 
energy). Such low energy $\alpha$ particles are stopped very rapidly, so the 
RTPC was designed to be as close as possible to the target and fit inside the 
230~mm diameter shell and cryostat wall of the solenoid magnet bore of CLAS.

@@

The main structural improvement compared to the 2005 RTPC~\cite{Fenker:2008zz} 
was to obtain a better support structure for the GEM foils. In the previous version,
the RTPC was essentially composed of two independent half-cylinder separated
by their own structures. In this design, the installation of the GEMs was not 
very practical and wrinkles were visible on the GEM surfaces. To allow for a 
better installation and, at the same time, keep the mechanical structure out of 
the drift volume, the present design is based on self supporting GEM cylinder.  
We realized these cylinders by using fiber glass rings glued to each end of the 
GEM foils in order to install them independently in the RTPC after gluing and 
soldering operations. The rigidity of the GEM foils was enough for the 
structure to be self-supporting and only the upstream end of the cylinder was 
fixed to the main mechanical support structure. This design only left a light 
fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

After the ionization, the released electrons drift to the cylindrical detection 
plane under the effect of the electric field. The electrons released close to 
the cathode take the most time to reach the readout pads, the cylindrical 
symmetry indicates that this maximum drift distance is the same for all tracks.  
We illustrate this in Figure~\ref{fig:RTPC_signals}, where we depict a typical 
$^{4}$He track. Due to the non-perfect experimental conditions, in particular 
possible contamination of our gas mixture, the drift speed changed during the 
three months long experimental run\footnote{We had to increase the gas flow 
   during the experiment due to a small leak in the RTPC, which concurred to 
   the large shift of drift time observed around run number 61600. While our 
   gas system was kept slightly over atmospheric pressure to limit 
   contamination from air or other external gases, it is likely that this leak 
   was the source of modification of the drift speed.}. This can be calibrated 
   by finding drift speed as function of experimental run-number. By measuring 
   a reference time, we can therefore infer the drift speed of the electrons in 
   the RTPC, by: $\textnormal{drift time} = \frac{\textnormal{drift path 
   length}}{\textnormal{drift speed}}$.

\begin{figure}[tb]
\centering
\includegraphics[{trim={0 12cm 0 0},clip, scale=0.4}]{RTPC_1_4.pdf}
\caption[]{A schematic drawing of a $^{4}$He track (in green) traversing the 
drift region, with the drift paths followed by the electrons (in blue). } 
\label{fig:RTPC_signals}
\end{figure}

Figure~\ref{fig:TDC_profile} presents the time profile of hits from identified 
tracks in one z-bin along the detector in one experimental run. We can clearly 
observe the dropping edge expected from geometrical considerations. In order to 
avoid the statistical fluctuations on finding the maximum drift time 
($T_{max}$) from the time profile, we define it in a more stable value, i.e., 
at which the dropping edge passes half the maximum number of hits in the 
histogram. This value was measured in bins along the 200~mm RTPC's length to 
take into account variations in the electric and magnetic field in the RTPC 
(see Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

\begin{figure}[t]
\centering
\includegraphics[scale=0.42]{hits_time_profile.pdf}
\caption[]{Time distribution of the hits in one z-bin along the RTPC in one 
experimental run. Trigger time is defined as T=15, the time unit is the ALTRO 
sampling time of 100~ns.} \label{fig:TDC_profile}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.42]{RunNumber_61452_TDCmax_Zslice.pdf}
\caption{Maximum time of collected hits as a function of the track
              position on the $z$-axis for one experimental run. } 
\label{fig:RunNumber_61551_TDCmax_Zslice}
\end{figure}

Figure~\ref{fig:Drift_run_number_1} shows the $T_{max}$ values for individual 
runs (approximately 2 hours long). We observe significant change in the drift 
speed before and after run 61600, while variations within these periods are 
around 2\%.

\begin{figure*}[t!]
\centering
\includegraphics[width=17.5cm]{Drift_run_number.pdf}
\caption{$T_{max}$ versus the experimental run numbers.} 
\label{fig:Drift_run_number_1}
\end{figure*}

@@

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.46]{FitResult_p2_11.pdf}
\caption{$\Delta \phi$ versus T distribution for tracks
from one bin in longitudinal position along the RTPC. The black line represents 
the final drift paths in this bin.}
\label{fig:DELTA_PHI_TDC}
\end{figure}

@@

We then compared this simulation to experiment on an event by event basis as 
shown in Figure~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated 
as the ratio of the measured ADCs to the simulated deposited energy. This 
provided our initial set of gains for each pads and it was applied to the 
experimental data. However, after this calibration some pads recorded lower ADCs 
than expected. So, we added a correction factor obtained by comparing  
the energy deposit on a given pad to its neighbours. To do so, we compared, for 
each track, the energy deposit on a pad to the energy deposit in the whole track. 
The gain correction factors are extracted by averaging this ratio for a sample 
of elastic events. Final results are shown in Figure~\ref{fig:dedx_p_exp_2nd}, 
where energy loss of particles is plotted against momentum over charge ratio. 
One can clearly see the band for $^4$He in its expected position\footnote{The 
   lower peak in the left module of the RTPC comes from an unidentified problem 
   in this half of the detector that concerns 7\% of all the elastic events.  
   Our best guess is that these events are linked with a high voltage supply 
   issue, that would sometimes bring the GEM gain down. These events pass all 
the elastic requirements and we found nothing that differentiates them from 
other events but their low recorded ADCs. For the calibration procedures, these 
events were therefore excluded.}.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Figure~\ref{fig:noise} for a particularly noisy 
channel. Its amplitude is similar to those of real tracks. About 18\% of the 
readout channels exhibit large contributions from this noise.  Due to its 
unique time-energy correlation, we use a pattern recognition algorithm and 
discard the hits coming from the channels that display this behavior on an 
event by event basis. The result of the procedure is illustrated in the bottom 
panel of Figure~\ref{fig:noise}.

@@

\subsection{Energy Loss Corrections}\label{sec_eloss}
Energy loss between the target and drift region was significant in the RTPC and 
necessitated a correction for optimal momentum reconstruction at the primary 
interaction vertex.  The dominant loss was in the 27~\textmu{}m thick Kapton target 
wall, with significant contributions also from the pressurized target gas and 
the foils before reaching the drift region.  Corrections were developed based 
on GEANT4 simulations with the full RTPC geometry and parameterized in terms of 
recoil curvature in the drift region and polar angle, separately for all recoil 
hypotheses ($p$, $d$, $^3$H, $^3$He, $^4$He).  At our average coherent $^4$He 
DVCS kinematics, energy losses were about 5 MeV, while for $e-^4$He elastic 
scattering losses were about 3 MeV, which corresponds to momentum corrections 
of 25\% and 15\%, respectively.
